---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# cumulative-incidence

The goal of the cumulative-incidence project is to provide a tutorial for JHS-HWG members who want to create cumulative incidence plots using `ggplot2`. To demonstrate the code in this repository, we use the `flchain` data from the `survival` package. These data are loaded and printed here.

This is a stratified random sample containing 1/2 of the subjects from a study of the relationship between serum free light chain (FLC) and mortality. The original sample contains samples on approximately 2/3 of the residents of Olmsted County aged 50 or greater.
```{r load-and-print-flchain, message = FALSE}

# the flchain data are in the survival package
library(survival)
# tidyverse is used for convenience
library(tidyverse)

# take a look at the data, 
# each column is printed with a few values to the right of its column name
glimpse(flchain)

```

# Create cumulative incidence data

The `cuminc()` function in the `cmprsk` package can generate data on the cumulative incidence of a censored outcome, accounting for competing risks if necessary.

```{r generate-cumulative-incidence}

# loading the cmprsk package to gain access to cuminc() function
library(cmprsk)

cml_inc = cuminc(
 # the first argument is the failure times, i.e., the time to event variable
 ftime   = flchain$futime, 
 # the second argument is the status variable (1 = event, 0 = censored)
 # (if you have competing risks, you will have values of 1, 2, 3, ... etc 
 #  for the different event types)
 fstatus = flchain$death,
 # the third argument is the group variable, which will be used to create 
 # one cumulative incidence curve per group.
 group   = flchain$creatinine > 1
)


```

Note here that we have 1350 observations as they had missing values for `creatinine`. You should be aware of this mechanic in case you have missing values in your grouping variable.


Next we use the `ggcompetingrisks` package to create a dataset with cumulative incidence estimates. `ggcompetingrisks` also makes plots, but they are a little harder to customize than a standard `ggplot2` plot.

```{r create-ggdat}

# load the survminer package to gain access to the
# ggcompetingrisks function 
library(survminer)

ggdat <- cml_inc %>%
  ggcompetingrisks(conf.int = T) %>%
  getElement('data') %>%
  as_tibble() %>%
  select(
    time,
    group, # same group as in cuminc(),
    est, # est is cumulative incidence,
    std # std is the standard deviation of est
  )

print(ggdat)

```




# one figure, two groups, no standard errors

This code will generate a figure that directly compares the incidence of an event between two or more groups. The code utilizes the data that we created using `cuminc()` above.

```{r}

ggplot(ggdat, aes(x = time, y = est, col = group))+
 geom_line(size=0.9)+
 # This is a theme that is consistent with many journals expectations
 # for figures
 theme_bw()+
 theme(
  panel.grid = element_blank(),
  text = element_text(size = 13, color = 'black', face = 'bold')
 ) +
 scale_y_continuous(label=scales::percent) +
 labs(x='Time, years', y = 'Cumulative Incidence of the event, %')

```






