---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# cumulative-incidence

The goal of the cumulative-incidence project is to provide a tutorial for JHS-HWG members who want to create cumulative incidence plots using `ggplot2`. To demonstrate the code in this repository, we use the `flchain` data from the `survival` package. These data are loaded and printed here.

```{r load-and-print-flchain, message = FALSE}

# the flchain data are in the survival package
library(survival)
# tidyverse is used for convenience
library(tidyverse)

# take a look at the data, 
# each column is printed with a few values to the right of its column name
glimpse(flchain)

```

# Create cumulative incidence data

The `cuminc()` function in the `cmprsk` package can generate data on the cumulative incidence of a censored outcome, accounting for competing risks if necessary.

```{r}

# loading the cmprsk package to gain access to cuminc() function
library(cmprsk)

cml_inc = cuminc(
 # the first argument is the failure times, i.e., the time to event variable
 ftime   = flchain$futime, 
 # the second argument is the status variable (1 = event, 0 = censored)
 # (if you have competing risks, you will have values of 1, 2, 3, ... etc 
 #  for the different event types)
 fstatus = flchain$death,
 # the third argument is the group variable, which will be used to create 
 # one cumulative incidence curve per group.
 group   = flchain$creatinine > 1
)
```

Note here that we have 1350 observations as they had missing values for `creatinine`. You should be aware of this mechanic in case you have missing values in your grouping variable.

